---
title: "Did the Formula 2022 Rule Change Improve Racing as Intended."
author: "by : Dimitri Rao, Eric Zhang, Ray Zang, Fergus Towler & Moses Weintraub"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r load-lib, include = FALSE}
library(tidyverse)
# Add any other libraries here



```


```{r load-data, include=FALSE}
# load your data here and any pre-processing/cleaning 
#  that you need for your report.
constructor_results <- read_csv('data/constructor_results.csv')
constructors <- read_csv('data/constructors.csv')
drivers <- read_csv('data/drivers.csv')
qualifying <- read_csv('data/qualifying.csv')
races <- read_csv('data/races.csv')
results <- read_csv('data/results.csv')
```
```{r Drivers and Constructors, message = FALSE, warning = FALSE }
new_drivers<-subset(drivers,select=-c(driverRef,code,dob, nationality, url))
new_drivers1<-subset(new_drivers,select=-c(number)) 
new_constructors<-subset(constructors,select=-c(constructorRef,nationality,url))
driver_data <- merge(qualifying, new_drivers1, by = "driverId")
constructor_data <- merge(qualifying, new_constructors, by = "constructorId")
qualifying_data <- merge(constructor_data, driver_data, by = "qualifyId")
qualifying_data <- qualifying_data[, -which(names(qualifying_data) == "constructorId.x")]
qualifying_data <- qualifying_data[, -which(names(qualifying_data) == "raceId.x")]
qualifying_data <- qualifying_data[, -which(names(qualifying_data) == "driverId.x")]
 qualifying_data <- qualifying_data[, -which(names(qualifying_data) == "number.x")]
qualifying_data <- qualifying_data[, -which(names(qualifying_data) == "position.x")]
qualifying_data <- qualifying_data[, -which(names(qualifying_data) == "q1.x")]
qualifying_data <- qualifying_data[, -which(names(qualifying_data) == "q2.x")]
qualifying_data <- qualifying_data[, -which(names(qualifying_data) == "q3.x")]
qd <- as.data.frame(qualifying_data)
colnames(qd) <- c("qualifyId", "name", "driverId", "raceId", "constructorId", "number", "position", "q1", "q2", "q3", "forename", "surname")
```

```{r constructorPointsPerSeasonTable, message = FALSE, warning = FALSE }
constructorPointsPerSeason<-left_join(constructors,constructor_results,by='constructorId')
season09<-constructorPointsPerSeason%>%
  filter(raceId>=1,raceId<=17)%>%
  group_by(name)%>%
  summarise(sum(points))
season10<-constructorPointsPerSeason%>%
  filter(raceId>=337,raceId<=355)%>%
  group_by(name)%>%
  summarise(sum(points))
season11<-constructorPointsPerSeason%>%
  filter(raceId>=841,raceId<=859)%>%
  group_by(name)%>%
  summarise(sum(points))
season12<-constructorPointsPerSeason%>%
  filter(raceId>=860,raceId<=879)%>%
  group_by(name)%>%
  summarise(sum(points))
season13<-constructorPointsPerSeason%>%
  filter(raceId>=880,raceId<=899)%>%
  group_by(name)%>%
  summarise(sum(points))
season14<-constructorPointsPerSeason%>%
  filter(raceId>=900,raceId<=918)%>%
  group_by(name)%>%
  summarise(sum(points))
season15<-constructorPointsPerSeason%>%
  filter(raceId>=926,raceId<=945)%>%
  group_by(name)%>%
  summarise(sum(points))
season16<-constructorPointsPerSeason%>%
  filter(raceId>=948,raceId<=968)%>%
  group_by(name)%>%
  summarise(sum(points))
season17<-constructorPointsPerSeason%>%
  filter(raceId>=969,raceId<=988)%>%
  group_by(name)%>%
  summarise(sum(points))
season18<-constructorPointsPerSeason%>%
  filter(raceId>=989,raceId<=1009)%>%
  group_by(name)%>%
  summarise(sum(points))
season19<-constructorPointsPerSeason%>%
  filter(raceId>=1010,raceId<=1030)%>%
  group_by(name)%>%
  summarise(sum(points))
season20<-constructorPointsPerSeason%>%
  filter(raceId>=1031,raceId<=1047)%>%
  group_by(name)%>%
  summarise(sum(points))
season21<-constructorPointsPerSeason%>%
  filter(raceId>=1052,raceId<=1073)%>%
  group_by(name)%>%
  summarise(sum(points))
season22<-constructorPointsPerSeason%>%
  filter(raceId>=1074,raceId<=1096)%>%
  group_by(name)%>%
  summarise(sum(points))
constructorPointsPerSeason<-full_join(season09,season10,by='name')
constructorPointsPerSeason<-full_join(constructorPointsPerSeason,season11,by='name')
constructorPointsPerSeason<-full_join(constructorPointsPerSeason,season12,by='name')
constructorPointsPerSeason<-full_join(constructorPointsPerSeason,season13,by='name')
constructorPointsPerSeason<-full_join(constructorPointsPerSeason,season14,by='name')
constructorPointsPerSeason<-full_join(constructorPointsPerSeason,season15,by='name')
constructorPointsPerSeason<-full_join(constructorPointsPerSeason,season16,by='name')
constructorPointsPerSeason<-full_join(constructorPointsPerSeason,season17,by='name')
constructorPointsPerSeason<-full_join(constructorPointsPerSeason,season18,by='name')
constructorPointsPerSeason<-full_join(constructorPointsPerSeason,season19,by='name')
constructorPointsPerSeason<-full_join(constructorPointsPerSeason,season20,by='name')
constructorPointsPerSeason<-full_join(constructorPointsPerSeason,season21,by='name')
constructorPointsPerSeason<-full_join(constructorPointsPerSeason,season22,by='name')
constructorPointsPerSeason <- constructorPointsPerSeason %>%
  rename(
    Constructor_Name = name,
    `2009` = `sum(points).x`,
    `2010` = `sum(points).y`,
    `2011` = `sum(points).x.x`,
    `2012` = `sum(points).y.y`,
    `2013` = `sum(points).x.x.x`,
    `2014` = `sum(points).y.y.y`,
    `2015` = `sum(points).x.x.x.x`,
    `2016` = `sum(points).y.y.y.y`,
    `2017` = `sum(points).x.x.x.x.x`,
    `2018` = `sum(points).y.y.y.y.y`,
    `2019` = `sum(points).x.x.x.x.x.x`,
    `2020` = `sum(points).y.y.y.y.y.y`,    
    `2021` = `sum(points).x.x.x.x.x.x.x`,
    `2022` = `sum(points).y.y.y.y.y.y.y`
  )
```
```{r driver championship, message = FALSE, warning = FALSE }
driverPointsPerSeason<-left_join(drivers,results,by='driverId')
Season09 <- driverPointsPerSeason %>% #season09
  filter(raceId >= 1, raceId <= 17) %>%
  group_by(driverId) %>%
  summarise(sum(points))
Season10 <- driverPointsPerSeason %>% #season10
  filter(raceId >= 337, raceId <= 355) %>%
  group_by(driverId) %>%
  summarise(sum(points))
Season11 <- driverPointsPerSeason %>% #season11
  filter(raceId >= 841, raceId <= 859) %>%
  group_by(driverId) %>%
  summarise(sum(points))
Season12 <- driverPointsPerSeason %>% #season12
  filter(raceId >= 860, raceId <= 879) %>%
  group_by(driverId) %>%
  summarise(sum(points))
Season13 <- driverPointsPerSeason %>% #season13
  filter(raceId >= 880, raceId <= 899) %>%
  group_by(driverId) %>%
  summarise(sum(points))
Season14 <- driverPointsPerSeason %>% #season14
  filter(raceId >= 900, raceId <= 918) %>%
  group_by(driverId) %>%
  summarise(sum(points))
Season15 <- driverPointsPerSeason %>% #season15
  filter(raceId >= 926, raceId <= 945) %>%
  group_by(driverId) %>%
  summarise(sum(points))
Season16 <- driverPointsPerSeason %>% #season16
  filter(raceId >= 948, raceId <= 968) %>%
  group_by(driverId) %>%
  summarise(sum(points))
Season17 <- driverPointsPerSeason %>% #season17
  filter(raceId >= 969, raceId <= 988) %>%
  group_by(driverId) %>%
  summarise(sum(points))
Season18 <- driverPointsPerSeason %>% #season18
  filter(raceId >= 989, raceId <= 1009) %>%
  group_by(driverId) %>%
  summarise(sum(points))
Season19 <- driverPointsPerSeason %>% #season19
  filter(raceId >= 1010, raceId <= 1030) %>%
  group_by(driverId) %>%
  summarise(sum(points))
Season20 <- driverPointsPerSeason %>% #season20
  filter(raceId >= 1031, raceId <= 1047) %>%
  group_by(driverId) %>%
  summarise(sum(points))
Season21 <- driverPointsPerSeason %>% #season21
  filter(raceId >= 1052, raceId <= 1073) %>%
  group_by(driverId) %>%
  summarise(sum(points))
Season22 <- driverPointsPerSeason %>% #season22
  filter(raceId >= 1074, raceId <= 1096) %>%
  group_by(driverId) %>%
  summarise(sum(points))
driverPointsPerSeason<-full_join(Season09,Season10,by='driverId')
driverPointsPerSeason<-full_join(driverPointsPerSeason,Season11,by='driverId')
driverPointsPerSeason<-full_join(driverPointsPerSeason,Season12,by='driverId')
driverPointsPerSeason<-full_join(driverPointsPerSeason,Season13,by='driverId')
driverPointsPerSeason<-full_join(driverPointsPerSeason,Season14,by='driverId')
driverPointsPerSeason<-full_join(driverPointsPerSeason,Season15,by='driverId')
driverPointsPerSeason<-full_join(driverPointsPerSeason,Season16,by='driverId')
driverPointsPerSeason<-full_join(driverPointsPerSeason,Season17,by='driverId')
driverPointsPerSeason<-full_join(driverPointsPerSeason,Season18,by='driverId')
driverPointsPerSeason<-full_join(driverPointsPerSeason,Season19,by='driverId')
driverPointsPerSeason<-full_join(driverPointsPerSeason,Season20,by='driverId')
driverPointsPerSeason<-full_join(driverPointsPerSeason,Season21,by='driverId')
driverPointsPerSeason<-full_join(driverPointsPerSeason,Season22,by='driverId')
driverPointsPerSeason <-driverPointsPerSeason %>%
  rename(
    Driver_ID = driverId,
    `2009` = `sum(points).x`,
    `2010` = `sum(points).y`,
    `2011` = `sum(points).x.x`,
    `2012` = `sum(points).y.y`,
    `2013` = `sum(points).x.x.x`,
    `2014` = `sum(points).y.y.y`,
    `2015` = `sum(points).x.x.x.x`,
    `2016` = `sum(points).y.y.y.y`,
    `2017` = `sum(points).x.x.x.x.x`,
    `2018` = `sum(points).y.y.y.y.y`,
    `2019` = `sum(points).x.x.x.x.x.x`,
    `2020` = `sum(points).y.y.y.y.y.y`,    
    `2021` = `sum(points).x.x.x.x.x.x.x`,
    `2022` = `sum(points).y.y.y.y.y.y.y`
  )

```
```{r quali split table, message = FALSE, warning = FALSE }
timeToSec <- function(time_str) {
  time_parts <- strsplit(time_str, ":")[[1]]
  minutes <- as.numeric(time_parts[1])
  seconds <- as.numeric(time_parts[2])
  total_seconds <- minutes * 60 + seconds
  return(total_seconds)
}
avg_quali <- qd %>%
  select(q3, raceId) %>%
  filter(
    raceId %in% 1:17 |
    raceId %in% 337:355 |
    raceId %in% 841:859 |
    raceId %in% 860:879 |
    raceId %in% 880:899 |
    raceId %in% 900:918 |
    raceId %in% 926:945 |
    raceId %in% 948:968 |
    raceId %in% 969:988 |
    raceId %in% 989:1009 |
    raceId %in% 1010:1030 |
    raceId %in% 1031:1047 |
    raceId %in% 1052:1073 |
    raceId %in% 1074:1096
  ) %>%
  mutate(q3 = sapply(q3, timeToSec)) %>%
  group_by(raceId) %>%
  summarise(q3_sd = sd(q3,na.rm=TRUE))

m09 <- avg_quali %>%
  filter(raceId %in% 1:17) %>%
  summarise(mean_sd_q3 = mean(q3_sd, na.rm = TRUE))
m10 <- avg_quali %>%
  filter(raceId %in% 337:355) %>%
  summarise(mean_sd_q3 = mean(q3_sd, na.rm = TRUE))
m11 <- avg_quali %>%
  filter(raceId %in% 841:859) %>%
  summarise(mean_sd_q3 = mean(q3_sd, na.rm = TRUE))
m12 <- avg_quali %>%
  filter(raceId %in% 860:879) %>%
  summarise(mean_sd_q3 = mean(q3_sd, na.rm = TRUE))
m13 <- avg_quali %>%
  filter(raceId %in% 880:899) %>%
  summarise(mean_sd_q3 = mean(q3_sd, na.rm = TRUE))
m14 <- avg_quali %>%
  filter(raceId %in% 900:918) %>%
  summarise(mean_sd_q3 = mean(q3_sd, na.rm = TRUE))
m15 <- avg_quali %>%
  filter(raceId %in% 926:945) %>%
  summarise(mean_sd_q3 = mean(q3_sd, na.rm = TRUE))
m16 <- avg_quali %>%
  filter(raceId %in% 948:968) %>%
  summarise(mean_sd_q3 = mean(q3_sd, na.rm = TRUE))
m17 <- avg_quali %>%
  filter(raceId %in% 968:988) %>%
  summarise(mean_sd_q3 = mean(q3_sd, na.rm = TRUE))
m18 <- avg_quali %>%
  filter(raceId %in% 989:1009) %>%
  summarise(mean_sd_q3 = mean(q3_sd, na.rm = TRUE))
m19 <- avg_quali %>%
  filter(raceId %in% 1010:1030) %>%
  summarise(mean_sd_q3 = mean(q3_sd, na.rm = TRUE))
m20 <- avg_quali %>%
  filter(raceId %in% 1031:1047) %>%
  summarise(mean_sd_q3 = mean(q3_sd, na.rm = TRUE))
m21 <- avg_quali %>%
  filter(raceId %in% 1052:1073) %>%
  summarise(mean_sd_q3 = mean(q3_sd, na.rm = TRUE))
m22 <- avg_quali %>%
  filter(raceId %in% 1074:1096) %>%
  summarise(mean_sd_q3 = mean(q3_sd, na.rm = TRUE))
result_df <- bind_rows(
  m09 %>% mutate(Year = 2009),
  m10 %>% mutate(Year = 2010),
  m11 %>% mutate(Year = 2011),
  m12 %>% mutate(Year = 2012),
  m13 %>% mutate(Year = 2013),
  m14 %>% mutate(Year = 2014),
  m15 %>% mutate(Year = 2015),
  m16 %>% mutate(Year = 2016),
  m17 %>% mutate(Year = 2017),
  m18 %>% mutate(Year = 2018),
  m19 %>% mutate(Year = 2019),
  m20 %>% mutate(Year = 2020),
  m21 %>% mutate(Year = 2021),
  m22 %>% mutate(Year = 2022)
)

```

## Research Question

For the 2022 season, Formula 1 overhauled its rules and regulations with regards to car design and also introduced a budget cap per constructor. The aim of these changes was to improve racing by making the cars have less complex and intricate bodies so that lessen their disruption of the airflow behind the car. Previously, the disrupted airflow would compromise the car behind as the aerodynamics on the car behind would be as effective. Typically, one effect of the disrupted airflow was that many races had few overtakes occur which in turn led to the races being boring for spectators. The cars from 2022 onward, however should be able to follow each other more closely. Additionally, the budget cap will reduce the performance difference which will further improve overall racing. 

Did all of these changes have their intended effect?  

Through data analysis, we will determine whether or not the new rules worked.



## Data

This is a large data set containing all information on the Formula 1 races, drivers, constructors, qualifying, circuits, lap times, pit stops, championships from 1950 till the latest 2023 season. From this data, we looked at the 2009-2022 seasons as 2009 was the begining of a new era with regards to car design. 

RaceId in results assigns a specific number to each race, this can be used to identify the season by using the groupings we made.
In the results table, milliseconds is the the time a driver took to finish a race.
DriverId in drivers assigns each driver a number.
Points from Results shows the points each driver received from each race.
q3 contains the qualifying time of each driver from the third qualifying session. q3 had to be converted from time in minutes into milliseconds.




## Findings

### Introduction ###
In October 2019, Formula 1 announced sweeping changes to its rules that intended to make racing more competitive, something the sport had been seriously lacking since 2017. These changes were meant to be implemented in 2021 but due to COVID-19 pandemic, they were pushed back to 2022. Rule changes included substantial changes to the design of the car as well as a budget cap for every team (the term “constructor” is interchangeably used to refer to team). For our analysis, which consisted of figuring out whether the 2022 Formula 1 rule changes made racing more competitive, we looked at data between the years of 2009 to 2022. In 2009, Formula 1 cars were in the middle of the V8 engine era and were beginning to become substantially more complex than ever before. We used 2022 as our final data point because during production of the project, the 2023 season was still ongoing.  
To analyse the impact a new car design and budget caps would have on making racing more competitive, we looked at qualifying times, race times, constructor championship points and driver championship points. This gave us an insight into the races and the total amount of points per season.  
  

### Qualifying Split Time ###
```{r quali split graph, message = FALSE, warning = FALSE }
ggplot(result_df, aes(x = Year, y = mean_sd_q3)) +
  geom_point() +
  geom_smooth(method = 'lm', se = FALSE) +
  labs(
    title = "Average qualifying split time per season",
    x = "Year",
    y = "Average qualifying split time in seconds",
    subtitle = '2009 - 2022'
  )


```

For the qualifying and race times, we found the average standard deviation of each season for the final race time and the times in qualifying session three (Q3) which showed us the average split between the field. The split is the time difference between two drivers. When looking at race data, we expect the split to be around five seconds, but in qualifying, where drivers race over a single lap to determine the starting order for the race, we expect the difference to be less than a second. To plot this information onto a data visualisation, we converted the time which was in the form HH:MM:SS, into a numeric value in seconds. Then, we created a function timeToSec which does this and plotted this against each year as a numeric value to show a line of best fit.   

For qualifying, we only focused on Q3 which contains the ten most competitive drivers out of the twenty racing each weekend. In qualifying, the overall performance of the car is more prominent than in the race. During the race, driver skill, team strategy, and luck all play a major role in determining the result. Because of this, we found the average standard deviation of the qualifying times per season then plotted this per season. Our graph shows a clear increase in the difference in qualifying time. This is true for 2022 with the value being higher than that of both 2014 and 2016, as one constructor, Mercedes, was highly dominant in 2014 and 2016.  

### Average Split time per season ###
``` {r Sum of Races Time Per Race, message = FALSE, warning = FALSE }
results %>%
  arrange(raceId,positionOrder)%>%
  mutate(
  split_time =  if_else(.$positionOrder == 1 ,0,
                        if_else(.$positionOrder == "//N" ,0, as.integer(.$milliseconds) - lag(as.integer(.$milliseconds)))),
  season = case_when(
    .$raceId >= 1 & .$raceId <= 17 ~ "2009",
    .$raceId >= 337 & .$raceId <= 355 ~ "2010",
    .$raceId >= 841 & .$raceId <= 859 ~ "2011",
    .$raceId >= 860 & .$raceId <= 879 ~ "2012",
    .$raceId >= 880 & .$raceId <= 899 ~ "2013",
    .$raceId >= 900 & .$raceId <= 918 ~ "2014",
    .$raceId >= 926 & .$raceId <= 945 ~ "2015",
    .$raceId >= 948 & .$raceId <= 968 ~ "2016",
    .$raceId >= 969 & .$raceId <= 988 ~ "2017",
    .$raceId >= 989  & .$raceId <= 1009 ~ "2018",
    .$raceId >= 1010 & .$raceId <= 1030 ~ "2019",
    .$raceId >= 1031  & .$raceId <= 1047 ~ "2020",
    .$raceId >= 1052 & .$raceId <= 1073 ~ "2021",
    .$raceId >= 1074 & .$raceId <= 1096 ~ "2022"
  )
  )%>%
  group_by(season)%>%
  filter(split_time !=0) %>%
  drop_na(season) %>%
  summarise(y =mean(split_time, na.rm = TRUE)) %>%
  ggplot() +
  geom_point(mapping = aes(x = season, y = y/1000))+
  labs(
    title = "Average Split Time per season",
    subtitle = "2009-2022",
    x = "Season",
    y= "Average Split time (seconds)"
  )

```

For race time, we looked at the mean split time at the end of each race, split these outcomes by season and graphed our results. The split time is the difference in time between the winning driver and the proceeding drivers. The split times were given in milliseconds in the data set, so all we had to do was find the winner by looking at the position order, exclude all non-recorded times, and run down the list extracting split times. To find the raceId's per season, we created a document called “General Info” and wrote the ranges of raceId for each season which we used throughout the document.   

While we were unable to make a line of best fit due to the points being scattered, key trends are still visible. First, we can see that 2011 has a very high average split time compared to the couple years preceding and following it. This is because one driver, Sebastian Vettel, had a very dominant season which is reflected in the high average split time. Between 2014-2017, the average split time is very high due to Mercedes having a car that was very dominant. It is important to note that in 2014, Formula 1 moved from a V8 engine to a V6 turbo-hybrid engine and Mercedes was clearly the most successful constructor out of the ten that compete each season in adapting to these regulation changes. 2017 has the highest average split time due to Mercedes’ dominant car and one of their drivers, Nico Rosberg, retiring after the 2016 season with his replacement being a lesser driver. From 2018 to 2021, the average split time decreased because another constructor, Red Bull, began to become more competitive with Mercedes and Red Bull finally matched Mercedes in 2021. 2020 is an outlier in our data and this is due to the COVID-19 pandemic having a substantial impact on how much constructors could develop their cars. In 2022, the race split decreased as the new aerodynamic regulations made it easier for cars to follow each other which decreased split times. However, a lower average split time does not directly correlate to an increase in overtakes, an overtake is when a car “overtakes” another car during the race, something the rule changes were meant to help increase. In 2022, overtakes did not increase as Red Bull had a car that was much superior to the rest of the field.  
  

### Standard Deviation of Constructor and driver points ###
```{r makingConstructorGraph, message = FALSE, warning = FALSE }
constructorPointsPerSeason %>%
  summarise(
    sd_2009 = sd((`2009`), na.rm = TRUE),
    sd_2010 = sd((`2010`) , na.rm = TRUE),
    sd_2011 = sd((`2011`), na.rm = TRUE),
    sd_2012 = sd((`2012`), na.rm = TRUE),
    sd_2013 = sd((`2013`), na.rm = TRUE),
    sd_2014 = sd((`2014`), na.rm = TRUE),
    sd_2015 = sd((`2015`), na.rm = TRUE),
    sd_2016 = sd((`2016`), na.rm = TRUE),
    sd_2017 = sd((`2017`), na.rm = TRUE),
    sd_2018 = sd((`2018`), na.rm = TRUE),
    sd_2019 = sd((`2019`), na.rm = TRUE),
    sd_2020 = sd((`2020`), na.rm = TRUE),
    sd_2021 = sd((`2021`), na.rm = TRUE),
    sd_2022 = sd((`2022`), na.rm = TRUE)
  ) %>%
  gather(key = "year", value = "sd") %>%
  mutate(year = as.numeric(str_extract(year, "\\d+"))) %>%
  ggplot(aes(x = year, y = sd)) +
  geom_point() +
  geom_smooth(method = 'lm', na.rm = TRUE, se = FALSE) +
  labs(
    title = "Standard Deviation of Points per Race for Each Season Constructor",
    x = "Season",
    y = "Standard Deviation",
    subtitle = '2009-2022'
  )
```

```{r Standard diviation of Drivers, message = FALSE, warning = FALSE }
driverPointsPerSeason%>%
  summarise(
    sd_2009 = sd((`2009`), na.rm = TRUE),
    sd_2010 = sd((`2010`), na.rm = TRUE),
    sd_2011 = sd((`2011`), na.rm = TRUE),
    sd_2012 = sd((`2012`), na.rm = TRUE),
    sd_2013 = sd((`2013`), na.rm = TRUE),
    sd_2014 = sd((`2014`), na.rm = TRUE),
    sd_2015 = sd((`2015`), na.rm = TRUE),
    sd_2016 = sd((`2016`), na.rm = TRUE),
    sd_2017 = sd((`2017`), na.rm = TRUE),
    sd_2018 = sd((`2018`), na.rm = TRUE),
    sd_2019 = sd((`2019`), na.rm = TRUE),
    sd_2020 = sd((`2020`), na.rm = TRUE),
    sd_2021 = sd((`2021`), na.rm = TRUE),
    sd_2022 = sd((`2022`), na.rm = TRUE)
  ) %>%
  gather(key = "year", value = "sd") %>%
  mutate(year = as.numeric(str_extract(year, "\\d+"))) %>%
  ggplot(aes(x = year, y = sd)) +
  geom_point() +
  geom_smooth(method = 'lm', na.rm = TRUE, se = FALSE) +
  labs(
    title = "Standard Deviation of Points per Race for Each Season Driver",
    x = "Season",
    y = "Standard Deviation",
    subtitle = '2009-2022'
  )

```

In the graphs above, we plotted the Standard Deviation of Points per Race for Each Season of each constructor and driver. When looking at the standard deviation of both constructor and driver points, 2009 is undoubtedly an anomaly with regards to the Standard Deviation of points. This is because in 2010, Formula 1 changed how it awarded points after each race. From 2010 onwards, points are awarded to the top ten finishers on a sliding scale from 25 to 1 while before 2010, points were only awarded to the top eight finishers on a scale from 10 to 1. The next major change with regards to standard deviation for the drivers' championships occurred in 2012 when Formula 1 experienced one of its most competitive years ever. With the top two drivers separated by just three points, 281 to 278, and the top six separated by only 102 points, the standard deviation of points was quite low. In 2014, the first season of the V6 turbo hybrid era, Mercedes was dominant scoring 701 points which was nearly double of the 405 points score by second place Red Bull. This dominance continued though the gap decreased in 2015, something that is visible on the constructor's graph. Following this blip, Mercedes continued its improvement until 2018 when constructors Red Bull and Ferrari began to make serious progress in trying to catch Mercedes. However, this only further differentiated these three constructors from the remaining seven leading to the standard deviation continuing to increase.  
  
Due to COVID-19, the 2020 season saw the standard deviation decrease in both graphs. COVID-19 impacted how much constructors could work on their cars and some constructors were more affected than others.  Furthermore, fewer races occurred in the 2020 season, so the top teams had fewer points to increase their gap. In the 2021 season, Red Bull and Mercedes were evenly matched opponents with both graphs showing 2021 to be below the line of best fit. In 2022, the first year of the new regulations, Red Bull built a dominant car which led to the standard deviation to increase in both graphs.  

### Conclusion ###
In conclusion, the regulation changes have not increased overall competitiveness though they have allowed individual races to become more competitive. The Q3 times trended upwards after the regulation changes which shows that budget caps have not reduced the difference in car performance. The split time analysis shows a reduction in 2022 relative to prior years which means that races in the 2022 season were closer. Therefore, the change in aerodynamic regulations has had the intended effect of allowing cars to follow each other more closely which has made races both closer and more competitive. Since the graphs that measure the standard deviation of constructor and driver points per race for each season both had a line of best fit considerably increase from 2009 to 2022, the regulation changes have not worked as intended. If the regulation changes were effective, the standard deviation would decrease. Furthermore, the trend of an increasing gap between high and low performing drivers and constructors has not been slowed. Overall, the dramatic changes in aerodynamics have allowed races to become closer but budget constraints have not stopped major constructors from continuing to dominate due to their superior infrastructure and machinery. As 2022 is the first year of the new Formula 1 cars, it is most certainly possible that it may just take more time for all the teams to understand the new regulations and increase their competitiveness.     





### References


Dataset:https://www.kaggle.com/datasets/rohanrao/formula-1-world-championship-1950-2020 (27/10/2023)
https://www.formula1.com/en/results.html/2012/drivers.html (29/11/2023)
