---
title: "Did the Formula 2022 Rule Change Improve Racing as Intended."
author: "by : Dimitri Rao, Eric Zhang, Ray Zang, Fergus Towler & Moses Weintraub"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r load-lib, include = FALSE}
library(tidyverse)
# Add any other libraries here



```


```{r load-data, include=FALSE}
# load your data here and any pre-processing/cleaning 
#  that you need for your report.
circuits <- read_csv("data/circuits.csv")
constructor_results <- read_csv('data/constructor_results.csv')
constructor_standings <- read_csv('data/constructor_standings.csv')
constructors <- read_csv('data/constructors.csv')
driver_standings <- read_csv('data/driver_standings.csv')
drivers <- read_csv('data/drivers.csv')
lap_times <- read_csv('data/lap_times.csv')
pit_stops <- read_csv('data/pit_stops.csv')
qualifying <- read_csv('data/qualifying.csv')
races <- read_csv('data/races.csv')
results <- read_csv('data/results.csv')
seasons <- read_csv('data/seasons.csv')
sprint_results <- read_csv('data/sprint_results.csv')
status <- read_csv('data/status.csv')



```


## Research Question

In 2022, Formula One completely changed the rules and regulations for the cars and introduce a budget cap. The aim being to improve the racing by making the cars have less complex and intricate bodies that disrupted the airflow behind the car. The disrupted airflow would compromise the car behind as the aerodynamics would not be able to act as effectively. This normally led to very few overtakes in boring races. But with these new cars, in theory, the cars can follow each other more closely and the budget cap will reduce the performance difference and improve racing overall. But did this work. Through data analysis, this group will determine whether these new rules worked.





## Data

This is a large data set containing all information on the Formula 1 races, drivers, constructors, qualifying, circuits, lap times, pit stops, championships from 1950 till the latest 2023 season.From this data, my group looked at 2009-2022 season as this was a much more modern era of F1. Older seasons would not help show an accurate . 

https://www.kaggle.com/datasets/rohanrao/formula-1-world-championship-1950-2020/data (Retrieved on 27/10/2023)




## Findings

The write-up of your project and findings go here. 

Think of this as the text of your presentation with some extra detail to cover what there was not time to discuss in the presentation. This might include any assumptions you made when doing your analysis, any limitations of the work you have done or any ideas you have for future work. Feel free to split this section into subsections to make it easier to read. 

The length should be roughly 1,500 words. 
If you want to use a word count addin, you can install this by copying and pasting the following into RStudio:

`devtools::install_github("benmarwick/wordcountaddin", type = "source", dependencies = TRUE)`

You will then need to restart RStudio. Once you have done that, select the text you want to count the words of, go to Addins, and select the `Word count` addin.
This addin counts words using two different algorithms, but the results should be similar and as long as you're in the ballpark of 1,500 words, you're good! 
The addin will ignore code chunks and only count the words in prose. If you don't want to use the addin you can always copy and paste the text into Microsoft Word to do a Word Count!

You can also load your data here and present any analysis results / plots.
Make sure to hide your code with  `echo = FALSE` unless the point you are trying to make is about the code itself, in which case you should show your code.





## References

List any references here. You should, at a minimum, list your data source.


